Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-12-03 leadership_across_cultures_contexts.Rmd
Several very minor edits.

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013], exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 (p. 58) notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain important leadership positions within the groups and represent the group to governmental forces. @aa01-062 (p. 173) describes that in these local 'clubs', "the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. @oc06-034 (p. 236) in describing kin group leaders among the Iban, small-scale highland horticulturalists primarily in Borneo and Sarawak, provides example of this suite of traits in leaders (which also include elements of prestige-based and charismatic leadership discussed above):

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 (p. 411) describes that, "If there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful categories. Variables were clustered together if they were positively correlated. Leader qualities formed two top-level clusters, which we identify as a *Material, social, and reproductive success* cluster, and a *Social orientation* cluster. See Figure \@ref(fig:cluster)A. Several of traits often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 (p. 47 ) describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

>Chiefs managed unusually large quantities of resources and it is agreed that the basis of such wealth was polygamy. Rank was quite directly expressed in the number of wives. The chief of Omarakana had 16 wives in Malinowski's time, while his predecessor Enamakala had 19 in 1893 (Annual Report 1893-4:19). In 1950 despite long-established missionary activity, there were still 13 (Powell 1960: 136). In traditional Trobriand society it seems that all leaders of village clusters of guyau status became polygamists. An emerging leader might also expect to be given wives by traditional allies outside the cluster while he might demand them from traditional rivals of subordinate subclans (Powell 1960:136)...

Most of the leader quality variables were found in the *Social orientation* cluster, which had two smaller and more weakly supported subclusters that we identified as *Ability to provide social benefits* and *Ability to impose social costs*. The social benefits subcluster contained the *Knowledgeable/intelligent* and *Experienced/accomplished* variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and these variables are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport), a point to which we shall return. As brief examples of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them." [fo07-056] (p. 85); and among the Highland Scots, es10-003 (p. 147) notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy in the context of cultural transitions. @nm09-056 (p. 183) describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:

>Despite her exceptional gifts -- a brilliant mind, beauty, selfconfidence, unusual oratorical abilities, and her educational attainments -- Kellogg is also the most controversial Iroquois leader of the twentieth century. It is clear from her many bizarre involvements that she misused her prodigious talents and/or was incapable of carrying out all the massive designs she had for her people's betterment. Although acknowledged today as a major force and brilliant person, she is accused by many Iroquois elders of swindling them out of hundreds of thousands of dollars in her abortive efforts to bring their land claims to fruition and of creating debilitating factionalism that impeded tribal development for decades. Unfortunately, because of her questionable ethies and her inability to carry out what she espoused, Kellogg is blamed today for all that went wrong in Iroquois history in the interwar period. Consequently, her life story had the feel of a Greek tragedy: she wanted to use her extraordinary abilities to help her people but ended up accused by them of being a common outlaw.

@@

The management cluster included many more variables, which were grouped into two subclusters that we identified as "Mediation" and "Organization". The mediation subcluster included variables such as *Resolve conflict* (the most widely supported function in our sample) and *Provide counsel* (also widely supported), which again are functions that are commonly noted in the anthropological literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 (p. 185) describes the nature of conflict resolution processes in the absence of strong authority:

@@

The larger organization subcluster included variables such as *Organize cooperation*, *Ritual functions*, and *Military command*. This subcluster aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 (p. 127) notes:

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership must also outweigh the costs. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits and social status, but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Our results cannot determine, however, if the benefits to leaders and followers outweighed their costs, but shed light on the social and biological pressures associated with leadership and social hierarchy. See Figure \@ref(fig:varsupportcb).

For both leaders and followers, in summary, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

@@

The strong support for material and social benefits associated with leadership (Figure \@ref(fig:varsupportcb)) and the colinearity of measures associated with material wealth, social capital, and reproductive success (Figure \@ref(fig:cluster)A), suggests processes of leadership and followership might be strong drivers of the emergence and intergenerational transmission of inequality [@bowles_emergence_2010; @mattison_evolution_2016]. Various sources of material wealth and strategic or institutionalized processes of nepotism are widespread cross-culturally, including among many societies often described as "egalitarian." Theses results suggests anthropologists and economists investigating the emergence and evolution of inequality, including material, social, and reproductive dimensions, should incorporate and pay close attention to group processes and social strategies associated with leadership and followership.

@@

This study is based on the ethnographic record. Our methods are designed to capture a broad range of ethnographic descriptions from a culturally representative, systematic sample. Nonetheless, these results are conditional upon the content ethnographers chose to discuss and the way they chose to discuss it. The ethnographic record is certainly an imperfect source of information on human cultural diversity and hence a limitation of the current results. We, however, view the existence of the ethnographic record and the availability and functionality of the eHRAF as an enormous benefit for comparative research, given many of the populations represented in these data no longer exist or live as they did during the time of much ethnographic description. Within these ethnographic descriptions there is a general male bias, which could reflect a male bias in community leadership or the more systemic male bias in the ethnographic record [see @rosaldo_woman_1974; @mukhopadhyay_anthropological_1988]. These data are also biased towards cultural models (rather than descriptions of observed behavior) and community leadership; descriptions of kin group and residential group leadership are less frequent. The high proportion of text records discussing political group leaders could be attributable to our broad search strategy or the content of the ethnographic record of "leadership" more generally.

>>>>

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013], exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain important leadership positions within the groups and represent the group to governmental forces. @aa01-062 [p. 173] describes that in these local 'clubs', "the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. @oc06-034 [p. 236] in describing kin group leaders among the Iban, small-scale highland horticulturalists primarily in Borneo and Sarawak, provides example of this suite of traits in leaders (which also include elements of prestige-based and charismatic leadership discussed above):

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "If there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful categories. Variables were clustered together if they were positively correlated. Leader qualities formed two top-level clusters, which we identify as a *Material, social, and reproductive success* cluster, and a *Social orientation* cluster. See Figure \@ref(fig:cluster)A. Several of traits often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

>Chiefs managed unusually large quantities of resources and it is agreed that the basis of such wealth was polygamy. Rank was quite directly expressed in the number of wives. The chief of Omarakana had 16 wives in Malinowski's time, while his predecessor Enamakala had 19 in 1893 (Annual Report 1893-4:19). In 1950 despite long-established missionary activity, there were still 13 (Powell 1960: 136). In traditional Trobriand society it seems that all leaders of village clusters of guyau status became polygamists. An emerging leader might also expect to be given wives by traditional allies outside the cluster while he might demand them from traditional rivals of subordinate subclans (Powell 1960:136)....

Most of the leader quality variables were found in the *Social orientation* cluster, which had two smaller and more weakly supported subclusters that we identified as *Ability to provide social benefits* and *Ability to impose social costs*. The social benefits subcluster contained the *Knowledgeable/intelligent* and *Experienced/accomplished* variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and these variables are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)), a point to which we shall return. As brief examples of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them" [@fo07-056, p. 85]; and among the Highland Scots, @es10-003 [p. 147] notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy in the context of cultural transitions. @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:

>Despite her exceptional gifts -- a brilliant mind, beauty, self confidence, unusual oratorical abilities, and her educational attainments -- Kellogg is also the most controversial Iroquois leader of the twentieth century. It is clear from her many bizarre involvements that she misused her prodigious talents and/or was incapable of carrying out all the massive designs she had for her people's betterment. Although acknowledged today as a major force and brilliant person, she is accused by many Iroquois elders of swindling them out of hundreds of thousands of dollars in her abortive efforts to bring their land claims to fruition and of creating debilitating factionalism that impeded tribal development for decades. Unfortunately, because of her questionable ethics and her inability to carry out what she espoused, Kellogg is blamed today for all that went wrong in Iroquois history in the interwar period. Consequently, her life story had the feel of a Greek tragedy: she wanted to use her extraordinary abilities to help her people but ended up accused by them of being a common outlaw.

@@

The management cluster included many more variables, which were grouped into two subclusters that we identified as "Mediation" and "Organization". The mediation subcluster included variables such as *Resolve conflict* (the most widely supported function in our sample) and *Provide counsel* (also widely supported), which again are functions that are commonly noted in the anthropological literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 [p. 185] describes the nature of conflict resolution processes in the absence of strong authority:

@@

The larger organization subcluster included variables such as *Organize cooperation*, *Ritual functions*, and *Military command*. This subcluster aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 [p. 127] notes:

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership must also outweigh the costs. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits and social status, but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

@@

The strong support for material and social benefits associated with leadership (Figure \@ref(fig:varsupportcb)) and the colinearity of measures associated with material wealth, social capital, and reproductive success (Figure \@ref(fig:cluster)A), suggests processes of leadership and followership might be strong drivers of the emergence and intergenerational transmission of inequality [@bowles_emergence_2010; @mattison_evolution_2016]. Various sources of material wealth and strategic or institutionalized processes of nepotism are widespread cross-culturally, including among many societies often described as "egalitarian." These results suggest anthropologists and economists investigating the emergence and evolution of inequality, including material, social, and reproductive dimensions, should incorporate and pay close attention to group processes and social strategies associated with leadership and followership.

@@

This study is based on the ethnographic record. Our methods are designed to capture a broad range of ethnographic descriptions from a culturally representative, systematic sample. Nonetheless, these results are conditional upon the content ethnographers chose to discuss and the way they chose to discuss it. The ethnographic record is certainly an imperfect source of information on human cultural diversity and hence a limitation of the current results. We nevertheless view the existence of the ethnographic record and the availability and functionality of the eHRAF as an enormous benefit for comparative research, given many of the populations represented in these data no longer exist or live as they did during the time of much ethnographic description. Within these ethnographic descriptions there is a general male bias, which could reflect a male bias in community leadership or the more systemic male bias in the ethnographic record [see @rosaldo_woman_1974; @mukhopadhyay_anthropological_1988]. These data are also biased towards cultural models (rather than descriptions of observed behavior) and community leadership; descriptions of kin group and residential group leadership are less frequent. The high proportion of text records discussing political group leaders could be attributable to our broad search strategy or the content of the ethnographic record of "leadership" more generally.
